---
title:  "Baseline Characteristics Tables with qwraps2"
date:   2016-12-19
tags: [R, data summaries, qwraps2]
---

Almost every biomedical research paper requires a "Table 1: baseline patient
characteristics."  Many developers have published tools to help streamline the
construction of such tables.  The `qwraps2::summary_table` function is my
contribution to the toolbox.

I have constructed hundreds of "Table 1"s while working as a biostatistics
consultant in a biostatistics department.  I've also constructed many 'helper
functions' to try and streamline the production of such tables.  However, every
project, every data set, every lead author, every target journal, etc., will
present slightly different requirements for the contents and formating of the
tables.  As such, functions which tried to "do it all" such as
`qwraps::tableone`, an exported function from my defunct R package, required
constant extension to provide the needed output from each nuanced project.

The [`tableone`](https://CRAN.R-project.org/package=tableone) package does a lot
of good things, and is a great tool for quickly building the baseline summary
tables.  It reminds me a lot of `qwraps::tableone`.  What my experiences have
taught me is that each row group, or even each row, might require some tweaks.
The approach to building the tables I've taken now is explicitly define the
summary statistic(s) I want for each variable in the data set, the formatting
for the summary statistic(s), and in a way that is easy to work with one or more
grouping variables.

Quick side note: `qwraps2` is a package of 'quick wrappers' or helper functions
I've found useful for formating results and generating some graphics.

We'll use the `mtcars` dataset for our examples.  Let's report several summary
statistics for miles per gallon, number of cylinders, and weight of the
vehicles.  The following summary is provided to illustrate the functions and
thus will include some summaries that would not be used in a publication.

Load and attach the qwraps2 namespace.  We'll set the `qwraps2_markup` option to
`markdown`.  If this option is not set, `qwraps2` uses
`get0ption(qwraps2_markup, "latex")` as the default markup language.

```{r}
library(qwraps2)
options(qwraps2_markup = 'markdown') # default is latex
```

The data summary we want will be 

* Miles Per Gallon
  * min
  * mean (sd)
  * median (iqr)
  * max
* Cylinders
  * mean
  * n (%) of four cylinders engines
  * n (%) of six cylinders engines
  * n (%) of eight cylinders engines
* Weight
  * range

For cylinders we'll report several things, the mean number of cylinders, and the
count (%) of 4, 6, and 8 cylinders cars.  In a publication we would likely not
report such a summary, treating cylinders as both a continuous and categorical
value.  However, doing so here helps to illustrate the flexibility of the
`summary_table` method.

Outlining the wanted summary statistics above as a list-of-lists will help
explain the construction of the summary object constructed below.  The
`summary_table` method takes two arguments, 

1. `.data`, a `data.frame` or a `grouped_df` object, and 
2. `summaries` a list-of-lists of right hand sided `formula`e defining the
   summary statistics.

The construction of the summary table is achieved via `dplyr::summarize_`.

The `mtcar_summaries` object constructed below, defines each needed row of the
summary table via a `formula`.  I've included the  `qwraps2` namespace for
clarity.

```{r}
mtcar_summaries <-
  list("Miles Per Gallon" =
       list("min:"         = ~ min(mpg),
            "mean (sd)"    = ~ qwraps2::mean_sd(mpg, denote_sd = "paren"),
            "median (IQR)" = ~ qwraps2::median_iqr(mpg),
            "max:"         = ~ max(mpg)),
       "Cylinders:" = 
       list("mean"             = ~ mean(cyl),
            "mean (formatted)" = ~ qwraps2::frmt(mean(cyl)),
            "4 cyl, n (%)"     = ~ qwraps2::n_perc0(cyl == 4),
            "6 cyl, n (%)"     = ~ qwraps2::n_perc0(cyl == 6),
            "8 cyl, n (%)"     = ~ qwraps2::n_perc0(cyl == 8)),
       "Weight" =
       list("Range" = ~ paste(range(wt), collapse = ", "))
       )
```

<br>


The table is constructed and printed with ease:
```{r, results = "asis"}
summary_table(mtcars, mtcar_summaries)
```

<br>


Extending the table to show the same summary by a grouping variable, we'll use
`am` (Transmission: 0 = automatic, 1 = manual), is done as follows:

```{r, results = "asis"}
summary_table(dplyr::group_by(mtcars, am), mtcar_summaries)
```

<br>

And lastly, building one table with a column for the whole data set and columns
for each transmission type is:
```{r, results = 'asis'}
cbind(summary_table(mtcars, mtcar_summaries),
      summary_table(dplyr::group_by(mtcars, am), mtcar_summaries))
```

<br>

Using `dplry::group_by` will alow you to build the table with more than one
grouping variable.  For example:
```{r, results = 'asis'}
cbind(summary_table(mtcars, mtcar_summaries),
      summary_table(dplyr::group_by(mtcars, am, vs), mtcar_summaries))
```

<br>


The formatting of the output is controlled by the
`qwraps2:::print.qwraps2_summary_table` and `qwraps2::qable` functions.

If you find bugs or have suggestions on how to extend and improve this tool
please create an [issue on github](https://github.com/dewittpe/qwraps2/issues).





