---
title:  "R: Updating Objects"
date:   2016-11-10
categories: [R, programming]
tags: [R, programming]
---

The `stats::update` function is one of my favorite tools in R.  Using this
function saves a lot of time and effort when needing to modify an object.
However, this function has its limits.  The following is an example of how to
extend the use of `stats::update`.

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

## Basics of the `stats::update` function
The
[documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/update.html)
for the `stats::update` function is a good starting place.  Of course, examples
are even better.  We'll use the `diamonds` data set from within the [ggplot2](
https://CRAN.R-project.org/package=ggplot2) package for the examples.

```{r ggplot2-data}
data("diamonds", package = "ggplot2")
dplyr::glimpse(diamonds)
```

A simple regression model for the price of diamonds as a function of `carat`,
`cut`, `color`, and `clarity`.


## Acknowledgements
I didn't figure out how to do this completely on my own.  I had posed a question
on [stackoverflow](http://stackoverflow.com/q/25272387/1104685) which was the
basis for this post and extensions.
